Efficient Unconstraining Parameter Transforms for Hamiltonian Monte Carlo

### 2022-07-20 tex/dummypaper.tex
make paper for simplex and add randomseed

Although it's possible to write HMC samplers that work for simply
constrained values such as upper- and/or lower-bounds
\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
far more challenging to do the same for complex constraints such as
simplices or positive definite matrices or for densities involving
multiple constrained values.  Instead, usually constrained values are mapped to unconstrained values before sampling
\cite{}.  A change of variables adjustment is made to the target density for these transforms to work. The tricky part is- there can be more than one suitable mapping for every constraint. This leads us to the issue of selecting a transform.

\section{Transforms}

To draw samples from a distribution $p_X(x)$, defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y}=\mathbb{R}^m$ for $m \ge n$.

Let $f\colon \mathcal{Y} \to \mathcal{X}$ be a smooth and continuous map.
Only when $m = n$ can $f$ be bijective.
When $m > n$, multiple values of $y \in \mathcal{Y}$ may map to a given value of $x = f(y)$.
In this case, to make the function bijective, we consider an additional smooth and continuous map $g\colon \mathcal{Y} \to \mathcal{Z}$, where $\mathcal{Z}$ can be uniquely parameterized by $m - n$ real degrees of freedom.
We then define $h\colon \mathcal{Y} \to \mathcal{X} \times \mathcal{Z}: y \mapsto (f(y), g(y)) = (x, z)$.
$f$ and $g$ must be chosen so that $h$ is bijective almost-everywhere;
that is, the values of $y$ for which $h$ is non-bijective must have no probability mass.
When $h$ is non-bijective at single points, we call these singularities.
While singularities themselves are in general not a problem for sampling with MCMC, the region of a target density near the singularity tends to have high curvature.

@@

If $X$ is a random variable with density $p_X$ and $Y = f(X)$ for a
smooth and bijective function $f$, then
\[
  p_Y(y) = p_X(f^{-1}(y)) \absdet{J_{f^{-1}}(y)},
\]
where the Jacobian of the inverse transform is defined by
\[
  J_{f^{-1}}(y) = \frac{\partial}{\partial y} \, f^{-1}(y)
\]
and
\[
  \absdet{J_{f^{-1}}(y)}
  = \abs{\det J_{f^{-1}}(y)}.
\]

\subsection{Target Density}
Given a proper joint distribution with a smooth and continuous density $p_{X,Y}(x, y | \theta)$ for some parameters $\theta$, we can use the usual change of variables formula to write

\[
  p_Y(y | \theta) = p_{X,Z}(f(y), g(y) | \theta) |J_h(y)|.
\]

To sample from a target density $p_X(x | \theta)$, let $p_{X,Z}(x, z | \theta) = p_X(x | \theta) p_Z(z | x, \theta)$.
Then we must choose some proper distribution $p_Z$ that is smooth and continuous.
Given that choice, we find that
\[
  p_Y(y | \theta) = p_X(f(y) | \theta) p_Z(g(y) | f(y), \theta) |J_h(y)|.
\]

Note that when $m > n$, it is insufficient to specify a transform $h$.
One must also select a prior $p_Z$, and it is desirable to establish some general properties that a given pair $h$ and $p_Z$ should have.
In addition to the above properties, the optimal pair would
1. discard as few degrees of freedom $|m - n|$ as possible.
2. produce a $p_Z$ and $|J_y|$ and their gradients that are efficient to compute.
3. produce a $p_Y$ that is easy to sample from.

We can expand a bit on the latter point.
As a general rule, the more normal $p_Y$ is, the easier it is to sample from (REF?).
Additionally, near singularities, small perturbations to $y$ tend to result in larger changes to $h(y)$ than far from the singularity.
This warps the volume around the singularity, which causes $|J_y(y)|$ to smoothly approach infinity near the singularity.
If $m$ is large, this is generally not a problem, as the "curse of dimensionality" works in our favor, and there is exponentially more volume away from the singularity than near the singularity.
However, for low $m$, it is necessary that $p_{X,Z}$ goes to 0 near the singularity faster than $|J_y(y)|$ goes to infinity.
While $p_X$ is generally fixed by the user-specified model and may not have this necessary property, the user may then choose a $p_Z$ that has this property.

\section{Unit simplex}

@@

Both of these corrections can be captured with the generalization
\[
  \mathrm{correction}
  = \pi(r) |J|
  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
\]
for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.

\subsubsection*{Acknowledgements}

>>>>

Although it's possible to write HMC samplers that work for simply
constrained values such as upper- and/or lower-bounds
\cite{neal2011mcmc} or unit vectors \cite{byrne2013geodesic}, it is
far more challenging to do the same for complex constraints such as
simplices or positive definite matrices or for densities involving
multiple constrained values.  Instead, usually constrained values are mapped to unconstrained values before sampling. These transform mappings are formed out of the constraints on the model parameters. A change of variables adjustment is made to the target density for these transforms to work. The tricky part is- there can be more than one suitable mapping for every constraint. This leads us to the issue of selecting a transform.

We evaluate the efficiency of unconstraining parameter transforms using metrics like effective sample size, leapfrog steps and squared error. Different transforms work well for different models, this paper aims to provide a useful analysis of the same, so practitioners can choose transforms that work well for their models.

\section{Transforms}
The transform from a constrained space $\mathcal{X}$ to an unconstrained space $\mathcal{Y}$ is a smooth and bijective function $f: \mathcal{X} \to \mathcal{Y}$. We aim to induce a density $h(y)$ on $\mathcal{Y}$ and define a smooth and continous map $g\colon \mathcal{Y} \to \mathcal{X}$, such that $h(y)$ is uniform over $\mathcal{X}$. 
%review and make this better
Uniformity is defined here to mean, if $y$, $y`$ $\sim$ $h(y)$,  then $g(y)$, $g(y`)$ $\sim$ uniform($\mathcal{X}$).

% check m,n relationship
%To draw samples from a $y$ $p_X(x)$, defined on a constrained space $\mathcal{X}$ that can be uniquely parameterized by $n$ real degrees of freedom, we instead perform sampling in an unconstrained space $\mathcal{Y}=\mathbb{R}^m$ for $m \ge n$.

%Only when $m = n$ can $f$ be bijective.
%When $m > n$, multiple values of $y \in \mathcal{Y}$ may map to a given value of $x = g(y)$.

%In this case, to make the function bijective, we consider an additional smooth and continuous map $g\colon \mathcal{Y} \to \mathcal{Z}$, where $\mathcal{Z}$ can be uniquely parameterized by $m - n$ real degrees of freedom.
%We then define $h\colon \mathcal{Y} \to \mathcal{X} \times \mathcal{Z}: y \mapsto (f(y), g(y)) = (x, z)$.
%$f$ and $g$ must be chosen so that $h$ is bijective almost-everywhere;
%that is, the values of $y$ for which $h$ is non-bijective must have no probability mass.
%When $h$ is non-bijective at single points, we call these singularities.
%While singularities themselves are in general not a problem for sampling with MCMC, the region of a target density near the singularity tends to have high curvature.

@@

If $Y \in \mathbb{R^N}$ is a random variable with density $p_Y(y)$, where $Y = g^{-1}(X)$, where $X \in \mathbb{R^M}$. Then for a smooth and bijective function $g$ which is the inverse transform from an unconstrained space to a constrained space, we can define its density as:
\[
  p_Y(y) = p_X(g(y)) \absdet{J_{g}(y)},
\]
where the Jacobian of the inverse transform is defined by
\[
  J_{g}(y) = \frac{\partial}{\partial y} \, g(y)
\]
and
\[
  \absdet{J_{g}(y)}
  = \abs{\det J_{g}(y)}.
\]
\section{Unit simplex}

@@

Both of these corrections can be captured with the generalization
\[
  \mathrm{correction}
  = \pi(r) |J|
  = r^n e^{-r^p/p} \prod_{i=1}^n x_i
  = \exp\left(\sum_{i=1}^n y_i - \frac{1}{p} \left(\sum_{i=1}^n e^{y_i}\right)^p\right),
\]
for $p > 0$, which corresponds to $r \sim \text{Generalized-Gamma}(1, n, p)$.
\subsection*{Acknowledgements}
